---
title: "Practice 2"
author: "Stefano Pellegrini"
date: "1/14/2020"
output: pdf_document
---

```{r load myData, include=FALSE}
load("exam.RData")
```

# Problem 1

## Ex 1.1 Compute the mean tooth length for all six combinations of supplement types and levels. Also provide the standard error of the mean for each situation.

```{r}
head(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

dose_0.5 <- ToothGrowth$dose == 0.5
dose_1.0 <- ToothGrowth$dose == 1.0
dose_2.0 <- ToothGrowth$dose == 2.0

supp_vc <- ToothGrowth$supp == "VC"
supp_oj <- ToothGrowth$supp == "OJ"

# supp VC
supp_0.5vc <- supp_vc & dose_0.5
supp_1.0vc <- supp_vc & dose_1.0
supp_2.0vc <- supp_vc & dose_2.0
# supp OJ
supp_0.5oj <- supp_oj & dose_0.5
supp_1.0oj <- supp_oj & dose_1.0
supp_2.0oj <- supp_oj & dose_2.0

# Split the dataframes
supp_0.5vc_df <- ToothGrowth[supp_0.5vc,]
supp_1.0vc_df <- ToothGrowth[supp_1.0vc,]
supp_2.0vc_df <- ToothGrowth[supp_2.0vc,]

supp_0.5oj_df <- ToothGrowth[supp_0.5oj,]
supp_1.0oj_df <- ToothGrowth[supp_1.0oj,]
supp_2.0oj_df <- ToothGrowth[supp_2.0oj,]

# list of df
list.df <- list(supp_0.5vc_df, supp_1.0vc_df, supp_2.0vc_df, 
             supp_0.5oj_df, supp_1.0oj_df, supp_2.0oj_df)
# get mean
mean.len <- sapply(list.df, function(x) mean(x$len))
# get se mean              # standard error of the mean = standard deviation of len / sqrt (n.data)
sd.len <- sapply(list.df, function(x) sd(x$len) / sqrt(nrow(x)))

data.frame(combinations = c("0.5 VC", "1 VC", "2 VC", 
                            "0.5 OJ", "1 OJ", "2 OJ"), 
           mean_len = mean.len, 
           mean_se = sd.len)

# or from Gherardo solution
combinations <- expand.grid(supp = levels(ToothGrowth$supp), dose = levels(ToothGrowth$dose))

temp <- apply(combinations, MARGIN = 1, function(x){ 
  ix <- ToothGrowth$supp == x[1] & ToothGrowth$dose == x[2]
return( c(mean = mean(ToothGrowth[ix, 1]), se = sd(ToothGrowth[ix, 1]) / sqrt(sum(ix) )))
} )

means <- cbind(combinations, t(temp) )
means
```

# Ex 1.2 We will investigate whether diﬀerent dose levels have the same eﬀect. Perform 0.05-level two sample t-tests with unequal variances to check whether to reject the following null hypotheses, and explain the result for each hypothesis: 

```{r}
t.test(supp_0.5oj_df$len, supp_1.0oj_df$len, var.equal = FALSE)

# or from gherardo
t.test(x = ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == "0.5", 1],
y = ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == "1", 1], var.equal = FALSE )
```


# Problem 3

## Ex 3.1 Implement an R function for the PDF of the Gaussian mixture distribution, Plot the PDF of GM(2,1,5,1,0.3). 

```{r}
dgaussmix <- function(x, mean1, sd1, mean2, sd2, w){
  stopifnot(w <= 1 && w >= 0)
  stopifnot(sd1 > 0 && sd2 > 0)
  w * dnorm(x, mean1, sd1) + (1 - w) * dnorm(x, mean2, sd2)
}
curve(dgaussmix(x, 2, 1, 5, 1, 0.3), from = -5, to = 10, main = "GM(2,1,5,1, 0.3)", ylab = "density")
```

## Ex 3.2 Estimate the ﬁve parameters of the Gaussian mixture using the 1000 observed longitude values. You can done this numerically in R with the optim function. Plot the ﬁtted Gaussian mixture on top of the histogram of the longitude data.

To obtain initial guess for the parameter of the Gaussian mixture for the longitude locations we plot the histogram

```{r}
hist(quakes$long, probability = TRUE, breaks = "FD")
```

We can divide the longitude location observations in two groups, before and after 175.

```{r}
before <- quakes$long[quakes$long < 175]
after <- quakes$long[quakes$long > 175]
```


We can now consider the following initial estimate for the Gaussian mixture:

```{r}
m1.init <- mean(before)        # Or just look at the histogram and try to think what the values of
sd1.init <- sd(before)         # the parameters can be
m2.init <- mean(after)
sd2.init <- sd(after)
w.init <- length(before) / length(after)
par.init <- c(m1.init, sd1.init, m2.init, sd2.init, w.init)
par.init
```


We define now the minus log-likelihood and then start the optimization

```{r}
gaussmix_mll <- function(par, data){
  if (par[5] > 1 || par[5] < 0 ){
    return(Inf)                         # why return inf?
  }
  if (par[2] < 0 || par[4] < 0){
    return(Inf)
  }
  -sum(log(dgaussmix(x = data, mean1 = par[1], sd1 = par[2], 
              mean2 = par[3], sd2 = par[4], w = par[5])))
}
par.est_gaussmix <- optim(f = gaussmix_mll, par = par.init, data = quakes$long)$par
par.est_gaussmix
```

I plot the density with estimated parameters.

```{r}
hist(quakes$long, probability = TRUE, breaks = "FD")
curve(dgaussmix(x, par.est_gaussmix[1], par.est_gaussmix[2], 
      par.est_gaussmix[3], par.est_gaussmix[4], par.est_gaussmix[5]), add = TRUE, col = "red")
```



## Ex 3.3 Consider now another model where the longitude locations are i.i.d. Gaussian distributed.

```{r}
# Methods of moments
mu = mean(quakes$long)
sigma = sd(quakes$long)

# Minus log likelihood
gauss_mll <- function(par, data){
  -sum(dnorm(x = data, mean = par[1], sd = par[2]))
} 

par.est_gauss <- optim(f = gauss_mll, par = c(mu, sigma),          # optim daesn't work in this the
                 data = quakes$long, control = list(maxit = 10000))# pars estimated by moments m.
```

```{r}
hist(quakes$long, probability = TRUE, breaks = "FD")
curve(dnorm(x, mean = mu, sd = sigma), add = TRUE, col = "red")
```

## Ex 3.4 Compute the AIC and BIC values for the simple Gaussian model and the Gaussian mixture model for the longitude data. Which model should be selected?


```{r}
mll.mixture <- gaussmix_mll(data = quakes$long, par = par.est_gaussmix)
mll.gauss <- (-sum(dnorm(x = quakes$long, mean = mu, sd = sigma, log = TRUE)))

# AIC
aic.mixture <- 2 * mll.mixture + 2 * length(par.est)
aic.gauss <- 2 * mll.gauss + 2 * 2
c(aic.mixture, aic.gauss)

# BIC
n <- nrow(quakes)
bic.mixture <- 2 * mll.mixture + length(par.est) * log(n)
bic.gauss <- 2 * mll.gauss + 2 * log(n)
c(mixture = bic.mixture, gauss = bic.gauss)
```

BIC and AIC scores indicates that the mixture model should be selected.

## Ex 3.5 Repeat the above ﬁtting procedure for the latitude and the depth data, and perform as usual model selection using AIC and BIC, which model should be used? 

### Latitude data fit

```{r}
hist(quakes$lat, probability = TRUE, breaks = "FD")
```

```{r}
# Parameters mixture
par.lat <- optim(par = c(-32, 7, -20, 7, 0.5), fn = gaussmix_mll, data = quakes$lat)$par

# Parameters gauss
mu.lat <- mean(quakes$lat)
sigma.lat <- sd(quakes$lat)

hist(quakes$lat, probability = TRUE, breaks = "FD")
curve(dgaussmix(x, par.lat[1], par.lat[2], par.lat[3], par.lat[4], par.lat[5]), 
                add = TRUE, col = "red")
curve(dnorm(x, mean = mu.lat, sd = sigma.lat), col = "blue", add = TRUE)
legend("topleft", legend = c("mixture", "gaussian"), col = c("red", "blue"), lty = 1)
```

```{r}
mll.mixture = gaussmix_mll(par.lat, data = quakes$lat)
mll.gauss = -sum(dnorm(quakes$lat, mu.lat, sigma.lat, log = TRUE))

# AIC
aic.mixture <- 2 * mll.mixture + 2 * length(par.lat)
aic.gauss <- 2 * mll.gauss + 2 * 2
c(mixture = aic.mixture, gauss = aic.gauss)

# BIC
n = nrow(quakes)
bic.mixture <- 2 * mll.mixture + length(par.lat) * log(n)
bic.gauss <- 2 * mll.gauss + 2 * log(n)
c(mixture = bic.mixture, gauss = bic.gauss)
```

The mixture model is to prefer (AIC and BIC).

### Latitude data fit

```{r}
hist(quakes$depth, probability = TRUE, breaks = 30)
```


```{r}
# Parameters mixture
par.depth <- optim(par = c(110, 100, 600, 100, 0.5), fn = gaussmix_mll, data = quakes$depth)$par

# Parameters gauss
mu.depth <- mean(quakes$depth)
sigma.depth <- sd(quakes$depth)

hist(quakes$depth, probability = TRUE, breaks = 30)
curve(dgaussmix(x, par.depth[1], par.depth[2], par.depth[3], par.depth[4], par.depth[5]), 
                add = TRUE, col = "red")
curve(dnorm(x, mean = mu.depth, sd = sigma.depth), col = "blue", add = TRUE)
legend("topleft", legend = c("mixture", "gaussian"), col = c("red", "blue"), lty = 1)
```

```{r}
mll.mixture = gaussmix_mll(par.depth, data = quakes$depth)
mll.gauss = -sum(dnorm(quakes$depth, mu.depth, sigma.depth, log = TRUE))

# AIC
aic.mixture <- 2 * mll.mixture + 2 * length(par.depth)
aic.gauss <- 2 * mll.gauss + 2 * 2
c(mixture = aic.mixture, gauss = aic.gauss)

# BIC
n = nrow(quakes)
bic.mixture <- 2 * mll.mixture + length(par.lat) * log(n)
bic.gauss <- 2 * mll.gauss + 2 * log(n)
c(mixture = bic.mixture, gauss = bic.gauss)
```

## 3.6 In this question we consider a generalized linear model with the log link and stations follows a Gaussian distribution.

```{r}
fit1 <- glm(stations ~ ., data = quakes, family = gaussian(link = "log"))
summary(fit1)
```

```{r}
fit2 <- glm(stations ~ . + I(mag^2), data = quakes, family = gaussian(link = "log"))
summary(fit2)
```

```{r}
fit3 <- glm(stations ~ . + I(mag^2), data = quakes, family = gaussian(link = "log"))
summary(fit2)
```

```{r}
fit3 <- glm(stations ~ lat + long + depth + exp(mag) + 
              I(exp(mag)^2), data = quakes, family = gaussian(link = "log"))
summary(fit3)

```

## Ex 3.7 Perform the log likelihood ratio test selection between model 1 and model 2.

```{r}
anova(fit1, fit2, test = "LRT")
```

The p-value is very small and we thus reject the null hypothesis (e.g. at a level 0.0005) that the simpler model fit1 is sufficient.

### Use instead AIC and BIC to perform model selection between model 1, model 2 and model 3.

```{r}
models <- list(fit1 = fit1, fit2 = fit2, fit3 = fit3)
sapply(models, function(m){
  c(AIC = AIC(m), BIC = BIC(m))
})

AIC(fit1, fit2, fit3)
BIC(fit1, fit2, fit3)
```

Model 2 is selected by both AIC and BIC.

## 3.8  Fit the Poisson regression models with the log link function.

```{r}
fit4 <- glm(stations ~ ., data = quakes, family = poisson(link = "log"))
summary(fit4)

fit5 <- glm(stations ~ . + I(mag^2), data = quakes, 
            family = poisson(link = "log"))
summary(fit5)

fit6 <- glm(stations ~ lat + long + depth + exp(mag) + I(exp(mag)^2), 
            data = quakes, family = poisson(link = "log"))
summary(fit6)
```

Perform model selection between model 4 and model 5 using the anova function. Perform model selection
between the three Poisson regression models using AIC and BIC.

```{r}
anova(fit4, fit5, test = "LRT")

models <- list(fit4 = fit4, fit5 = fit5, fit6 = fit6) 
sapply(models, function(m){
  c(AIC = AIC(m), BIC = BIC(m))
})
```

Model 5 (fit5) is selected by both AIC and BIC.

```{r}
fit7 <- glm(stations ~ . + I(mag^2), data = quakes, family = Gamma(link = "inverse" ))
summary(fit7)
?family
```



